---
title: "Project Storm"
author: "Tejas"
date: "June 6, 2019"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE, echo= TRUE)
```
# **Assessment of Weather Event Impact on the United States**
  
## Data Processing

Data used for this analysis were taken from the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. Within the weather data, which was taken between 1950-November 11th 2011, there were 902297 observation with 37 variables. 

```{r}
setwd("~/R Material/Coursera/Course 5 Reproducible Research/Week 4/Project Storm/repdata_data_StormData.csv")
WeatherDatabase <- read.csv("repdata_data_StormData.csv", stringsAsFactors = FALSE)
```

To address questions on health and economics, columns were selected based on their usefulness in the analysis. The following variables were selected:

1. STATE -> information of location of weather event
2. BGN_DATE,BGN_TIME,END_DATE,END_TIME -> information on time/duration of weather event
3. EVTYPE -> information on type of weather event
4. FATALITIES, INJURIES -> measurable variables of health
5. PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP -> measurable variables of economics

Using the *dplyr* **select()** function, important columns were retained in a new data table.

```{r}
library(dplyr)
#list of variables important to the analysis
UsefulColumns <- c("EVTYPE","BGN_DATE","FATALITIES", "INJURIES", "PROPDMG",
                   "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")
#Retain only the useful variables
WeatherDatabase <- select(WeatherDatabase, UsefulColumns)

```

In 1950, only data on Tornados was recorded. As years went by, more different types of weather events were recorded. It was only in January 1996 when all the unique weather events in this dataset were available to be recorded. Because we want to compare impacts between each weather event, we will focus our attention to events that occurred on or after Jan 1st 1996. 

The *dplyr* **filter()** function was utilized to narrow down the time period within the data table.
```{r}
#First, change BGN_DATE from a character class to a Date Class
WeatherDatabase$BGN_DATE <- as.Date(WeatherDatabase$BGN_DATE, format = "%m/%d/%Y")
WeatherDatabase <- filter(WeatherDatabase, BGN_DATE >= "1996-01-01")
head(WeatherDatabase)
#change all EVTYPE entries to upper case format, and then remove any event type containing
#summary
WeatherDatabase$EVTYPE <- toupper(WeatherDatabase$EVTYPE)
WeatherDatabase <- WeatherDatabase[-grep("^SUMMARY", WeatherDatabase$EVTYPE),]
```
###Cleaning EVTYPE

The event type variables causes a lot of issues for the analysis. There are only 48 different types of events specified by the Storm Data Documentation, yet there over 500 different event types in the actual data; caused by typos, or entries that don't follow the documentation guidelines.

I created a function that manually cleans through and compresses the number of different variables. For simplicity, event types like "heavy wind" and "strong wind" were collapsed into a single variable: Wind. After cleaning, only 35 different event types were available.
```{r}
#EVTYPE Cleaning
EVClean <- function(x){
      print(paste("Previous Unique Events:", length(unique(x))))
      x <- toupper(x)
      x <- trimws(x, which = c("both", "left", "right"))
      x <- gsub("BLIZZARD SUMMARY", "BLIZZARD", x)
      x <- gsub("COASTAL.*", "COASTAL FLOOD",x)
      x <- gsub(".*DUST DEV.*", "DUST DEVIL", x)
      x <- gsub(".*SMOKE*.","DENSE SMOKE", x)
      x <- gsub(".*DUST$", "DUST STORM", x)
      x <- gsub("^TSTM.*", "THUNDERSTORM WIND", x) # need to account for thunderstorm wind
      x <- gsub("HURRICANE.*", "HURRICANE", x)
      x[grepl("TYPHOON",x)] <- "HURRICANE"
      x <- gsub("^VOLCANIC.*", "VOLCANIC ASH", x)
      x[grepl("MARINE\\s[T]", x)] <- "MARINE THUNDERSTORM WIND"
      x[grepl("(?<!THUNDERSTORM\\s)(WIND)", x, perl = TRUE)] <- "WIND"
      x[grepl("SURF", x)] <- "HIGH SURF"
      x[grepl("FUNNEL",x)] <- "FUNNEL CLOUD"
      x[grepl("(?<=STORM\\s)(WIND)", x, perl = TRUE)] <- "THUNDERSTORM WIND"
      x[grepl("FLOOD", x)] <- "FLOODING"
      x[grepl("DEBRIS",x)] <- "DEBRIS FLOW"
      x[grepl("(?<!FREEZING\\s)(FOG)",x, perl= TRUE)] <- "DENSE FOG"
      x[grepl("HEAT", x)]<- "HEAT"
      x[grepl("HOT", x)]<- "HEAT"
      x[grepl("WARM", x)]<- "HEAT"
      x[grepl("HAIL",x)] <- "HAIL"
      x[grepl("ICE",x)] <- "ICE STORM"
      x[grepl("RIP",x)] <- "RIP CURRENT"
      x[grepl("SLEE",x)] <- "SLEET"
      x[grepl("WATERSPOUT",x)] <- "WATERSPOUT"
      x[grepl("FIRE",x)] <- "WILDFIRE"
      x[grepl("LAKE",x)] <- "LAKE-EFFECT SNOW"
      x[grepl("(?<!LAKE-EFFECT\\s)(SNOW)", x, perl = TRUE)] <- "HEAVY SNOW" 
      x[grepl("(?!.*STORM)(WINT)", x, perl= TRUE)] <- "WINTER WEATHER"
      x[grepl("COLD",x)] <- "EXCESSIVE COLD"
      x[grepl("COOL",x)] <- "EXCESSIVE COLD"
      x[grepl("DRY",x)] <- "DROUGHT"
      x[grepl("DRIEST",x)] <- "DROUGHT"
      
      print(paste("Unique Events:", length(unique(x))))
      return(x)
      
}
WeatherDatabase$EVTYPE <- EVClean(WeatherDatabase$EVTYPE)
```

There were still over 100 unique event types in the dataset. At this point, those event types were too difficult to sort into a category so rows not containing the 35 possible event types were removed entirely from the dataset.
```{r}
#Viable Event Names
EventNames <- toupper(c("Astronomical Low Tide", "Avalanche", "Blizzard","Coastal Flood","Debris Flow", "Dense Smoke", "Drought", "Dust Devil", "Dust Storm", "Excessive Heat", "Flash Flood", "Flood", "Frost/Freeze", "Funnel Cloud", "Hail", "Heat", "Heavy Rain","Heavy Snow", "High Surf", "Wind", "Hurricane", "Ice Storm", "Lake-Effect Snow", "Flooding", "Lightning", "Marine Thunderstorm Wind", "Rip Current", "Seiche",
"Sleet", "Storm Surge/Tide", "Thunderstorm Wind", "Tornado","Tropical Depression",
"Tropical Storm", "Tsunami", "Volcanic Ash", "Waterspout", "Wildfire", "Winter Storm",
"Winter Weather"))

RemovalIndex = integer()
print(paste("Before:", as.character(length(WeatherDatabase$EVTYPE))))

for (i in c(1:length(WeatherDatabase$EVTYPE))){
      for (x in c(1:length(EventNames))){
            if (WeatherDatabase$EVTYPE[i] == EventNames[x]){
                  break
            } else if (x == length(EventNames)) {
                  RemovalIndex = c(RemovalIndex, i)
                  break
                 
            }
      }
}

WeatherDatabase <- WeatherDatabase[-RemovalIndex,]
print(unique(WeatherDatabase$EVTYPE))
```

###Cleaning the Crop and Property Damage columns

To calculate the damages to property we must combine PROPDMG, which is a damage numerical, with its corresponding PROPDMGEXP value, which denotes a multiplier. Here is a list of multipliers:
1. H or h - 10^3
2. M or m - 10^6
3. B or b - 10^9
4. numerical x that ranges from 0-8 - 10^x
5. "?","+", "-" represent uncertainties and will be ignored for this analysis

```{r}
#First we create a function that will deal with Exp values and conver them into numbers
ExpConvert <- function(x) {
      if (grepl("K", x)== TRUE){
            x <- 1000
      } else if (grepl("M",x)== TRUE){
            x <- 10^6
      } else if (grepl("B", x)== TRUE){
            x <- 10^9
      } else if (grepl("H", x)== TRUE){
            x <- 100
      } else if (grepl("[012345678]", x)== TRUE){
            x <- (10^(as.numeric(x)))
      } else {
            x <- NA
      }
      return(x)
}

#Use the ExpConvert function on the entirety of the PROPDMGEXP and CROPDMGEXP columns

WeatherDatabase$PROPDMGEXP <- sapply(toupper(WeatherDatabase$PROPDMGEXP), ExpConvert)
WeatherDatabase$CROPDMGEXP <- sapply(toupper(WeatherDatabase$CROPDMGEXP), ExpConvert)

#Created a function that multiplies two column values for every row
DmgCombine <- function(Dmg, Exp){
      
      for (i in c(1:length(Dmg))) {
          Dmg[i] <- Dmg[i]*Exp[i] 
             
      }
      return (Dmg)
}

#Use DmgCombine to calculate both crop damage and property damage
WeatherDatabase <- mutate(WeatherDatabase, PROPDMG = DmgCombine(PROPDMG,PROPDMGEXP))
WeatherDatabase <- mutate(WeatherDatabase, CROPDMG = DmgCombine(CROPDMG,CROPDMGEXP))
head(WeatherDatabase$PROPDMG)
head(WeatherDatabase$CROPDMG)
```

Because there were so many event types, I decided to focus in on the top 5 events for each of the four categories: Fatalities, Injuries, Crop Damage, and Property Damage. My goal was to look at how these measured variables changed over time. To accomplish this, a function was created to rank and extract the top 5 five events for a given category.But, first NA values had to be removed so four categorical frames were created. 

```{r}
#Removes NA Values from a given dataframe
CheckNA <- function(NAFrame){
      RemovedNA <- NAFrame[!is.na(NAFrame[,3]),]
      return(RemovedNA)
}

FatalityFrame <- CheckNA(select(WeatherDatabase, EVTYPE:FATALITIES))
InjuryFrame <- CheckNA(select(WeatherDatabase, c(EVTYPE, BGN_DATE, INJURIES)))
PropertyFrame <- CheckNA(select(WeatherDatabase, c(EVTYPE, BGN_DATE, PROPDMG)))
CropFrame <- CheckNA(select(WeatherDatabase, c(EVTYPE, BGN_DATE, CROPDMG)))

#created alternate frames that computed the totals for the different event types
FatalitySumFrame <- group_by(FatalityFrame, EVTYPE) %>% 
     summarise(sum(FATALITIES))
head(FatalitySumFrame)
InjurySumFrame <- group_by(InjuryFrame, EVTYPE) %>% 
      summarise(sum(INJURIES))
PropertySumFrame <- group_by(PropertyFrame, EVTYPE) %>% 
      summarise(sum(PROPDMG))
CropSumFrame <- group_by(CropFrame, EVTYPE) %>% 
      summarise(sum(CROPDMG))

names(FatalitySumFrame) <- c("EVTYPE","FATALITIES")
names(InjurySumFrame) <- c("EVTYPE","INJURIES")
names(PropertySumFrame) <- c("EVTYPE","PROPDMG")
names(CropSumFrame) <- c("EVTYPE","CROPDMG")



#Function that finds the top 5 for one of the parameter categories: fatalities, injuries, cropdmg, propdmg
TopFive <- function(Frame, Category){
      Frame <- Frame[order(-Frame[,2]),]
      print(head(Frame))
      return(Frame[1:5,1])
}

#find top five values for each category
TopFatalities <- TopFive(FatalitySumFrame, "FATALITIES")$EVTYPE
TopInjuries <- TopFive(InjurySumFrame, "INJURIES")$EVTYPE
TopPropertyDmg <- TopFive(PropertySumFrame, "PROPDMG")$EVTYPE
TopCropDmg <- TopFive(CropSumFrame, "CROPDMG")$EVTYPE
```